Biotic interactions structure zooplankton metacommunity dynamics following a summer heatwave

### 2024-02-15 writing/Manuscript.Rmd
Updated based on edits from Mary

\newpage
# Abstract
Despite the key role of biotic interactions in structuring ecological communities, their influence is often overlooked in predictions of how communities respond to environmental change. Here, we present an experiment that tests hypotheses based on metacommunity theory about how abiotic responses, biotic interactions, and dispersal jointly determine the response of ecological communities to environmental perturbations. We established experimental zooplankton metacommunities across spatial temperature gradients, connected by three levels of dispersal, that experienced natural temporal variation in ambient temperature. Prior to a mid-summer heatwave, community composition varied across the spatial temperature gradients. The heatwave homogenized the metacommunities and when conditions cooled, communities diverged into multiple compositional states that were not associated with temperature. These states appear to have been driven by biotic interactions that prevented the reestablishment of the pre-heatwave thermal compositional gradients. This highlights how biotic interactions can prevent metacommunities from tracking temperature changes via dispersal-facilitated species sorting.

@@

As the climate changes, extreme weather events, such as heatwaves, droughts, and storms, are increasing in frequency and intensity [@ipcc2021], placing ecological communities under considerable stress [@hesketh2023]. To design strategies aimed at mitigating the impacts of climate change on natural ecosystems, we require a sound understanding of the processes that determine how species and ecological communities respond to environmental change. According to metacommunity theory, the response of communities to environmental change should depend on the interplay between three general processes: density-independent population responses to abiotic conditions, density-dependent biotic interactions (i.e., interactions between species), and dispersal between communities [@thompson2020]. There is considerable experimental evidence that each of these processes can play a key role in governing community responses [summarized in @leibold2018]. Yet, we still lack clear empirical examples to help understand how abiotic responses, biotic interactions, and spatial processes interactively determine metacommunity dynamics when conditions change.

@@

Metacommunity theory [@loreau2003; @thompson2020], and evidence from a series of experiments [@thompson2012; @symons2013; @limberger2019], suggests that dispersal can play an important role in facilitating community recovery following periods of extreme environmental conditions. Dispersal can allow species to re-establish in habitats in which they have been extirpated, acting to maintain biodiversity and ecosystem functioning [@loreau2003; @thompson2020]. Thus, dispersal promotes species sorting---that is, when species distributions and abundances are matched to the local conditions across space and time [@leibold2004]. Dispersal can also introduce species that are more tolerant to extreme conditions and if these species perform similar roles to the species they replace, they can compensate for their loss [@thompson2012; @symons2013; @limberger2019]. Colonizing species, however, can also greatly alter the composition and functioning of ecosystems, often with profoundly negative impacts, as highlighted by numerous examples of invasive species [@ricciardi2013]. In addition to facilitating colonization, dispersal can also act to homogenize communities across space, a process known as mass effects [@mouquet2003; @leibold2004]. If rates are high enough, dispersal can act to maintain populations in conditions in which they would otherwise not persist, or in which they would otherwise persist at lower abundances [@holt1985]. Therefore, environmental conditions are expected to have less influence on the composition of communities when dispersal rates are high, due to the homogenizing effects of dispersal [@loreau2003]. The overall expectation in metacommunity theory is that intermediate rates of dispersal should facilitate colonization and potentially community recovery from environmental change via species sorting, and high rates of dispersal should reduce the impact of environmental change on ecological communities through mass effects [@loreau2003; @thompson2020].

Biotic interactions are expected to cause context dependency in how the composition of communities changes with environmental conditions. This is because the abundance and distribution of any given species depends both on environmental conditions as well as on the abundance and distribution of the species with which it interacts [@hutchinson1957]. Competition and predation, for example, may prevent a population from establishing and growing when environmental conditions would otherwise be suitable [@gilman2010; @brown2014]. Thus, biotic interactions are a primary reason why communities that experience the same environmental conditions may differ in composition [@davis1998; @ives2004; @thompson2021]. Biotic interactions are a critical driver in successional dynamics that are common in forest communities [@clements1916]. In freshwater temperate lakes and ponds, seasonal succession in plankton dynamics is driven by an interplay between temperature, competition for nutrients and light, and food web dynamics, such that the composition of the community at any given time is not purely predictable based on density-independent abiotic conditions [@sommer2012]. Despite this, most classic metacommunity theory [e.g., @leibold2004; @loreau2003] assumes that competition is equal amongst all individuals and species, which precludes context dependence. More recent theory highlights the importance of biotic interactions, and makes the prediction that biotic interactions should reduce species sorting and lead to non-random patterns of species co-occurrence across space and time that are unrelated to local environmental conditions [@thompson2020]. When communities experience periods of extreme conditions, such as during a heatwave, experimental evidence suggests that biotic interactions may prevent the community from returning to its initial compositional state [@davis1998, @thompson2021].

Here we present the results of a replicated pond zooplankton metacommunity experiment, where we assess how abiotic responses, dispersal, and biotic interactions combine to shape the response and recovery of the community to a mid-summer heatwave. We test the following three hypotheses: H1 -- If community composition is primarily determined by environmental conditions (here, temperature), then species abundances should vary strongly across spatial and temporal temperature gradients and community composition should be predictable based on temperature at any point in time. H2 -- If intermediate rates of dispersal facilitate species sorting, then species' environmental responses should be strongest when dispersal is present, but not so high as to homogenize communities across space via mass effects and flatten environmental responses. H3 -- If biotic interactions impede species sorting or result in successional dynamics, then community composition should not be predictable based on environmental conditions alone and instead species will show strong negative and positive co-occurrence patterns over space and time.

@@

## Experimental design
The experiment consisted of a total of 48--1136L aquatic mesocosms (Rubbermaid Stock Tanks, Rubbermaid USA). Mesocosms were set up in groups of four, each group comprising a single metacommunity, for twelve metacommunities in total. Within each metacommunity, we established a spatial temperature gradient using aquarium heaters (JÄGER TruTsemp, EHEIM GmbH & Co KG, Germany) of different wattages: 100W, 200W, 300W, in addition to a treatment with no heater (i.e., 0W). Thermostats on the heaters were set to their maximum level, 37 \degree C, to ensure that they remained on continuously over the course of the entire experiment, delivering a consistent amount of heat. Mesocosms receiving heaters were placed on 5.08 cm thick rigid panel Styrofoam insulation to increase heat retention. The heaters brought the mesocosms a relatively consistent amount above ambient temperature, resulting in spatial differences in temperature that were maintained as the ambient temperatures fluctuated diurnally and over the course of the 12-week experiment (Figure \@ref(fig:temperature)). The experiment started on June 14, 2018, when the dispersal treatment was applied for the first time. The experiment lasted 12 weeks, ending on September 6, 2018. Samples for estimating zooplankton abundance and composition were collected on week four, eight, and at the end (week 12) of the experiment. Details on the establishment of the biological communities and the sampling protocol are provided in the Supplementary Materials.

Our experimental treatments resulted in a mean temperature difference of 2.95 \degree C ( $\pm 0.41$ \degree C SD) between the mesocosms without heaters and the mesocosms that received 300 W heaters (Figure \@ref(fig:temperature)). Ambient temperatures were moderate over the first five weeks of the experiment (until mid-July), with an average mean (max) daily temperature in the unheated mesocosms of `r all_temperature %>% filter(time < 6, wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time < 6, wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C; Figure \@ref(fig:temperature)). The warmest period of the experiment occurred from weeks six to eight (late July and early August), which resulted in a averaged mean (max) daily ambient temperatures of `r all_temperature %>% filter(time %in% c(6:8), wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% c(6:8), wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) and `r all_temperature %>% filter(time %in% c(6:8), wattage == 300) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% c(6:8), wattage == 300) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) in the 300W treatments. Ambient temperatures dropped following this warm period, continuing to decline over the final four weeks of the experiment, so that mean (max) daily temperatures in the unheated mesocosms averaged `r all_temperature %>% filter(time %in% max(time), wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% max(time), wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) in the final week.

Dispersal---none, low, or high---was applied within each metacommunity through reciprocal exchange of water and biota. For all dispersal rates, 3 x 10 L of water and biota were collected from each mesocosm using a 3L Van Dorn integrated sampler and placed into 3 - 19 L buckets. For the high dispersal treatment, one of these buckets was poured reciprocally into each of the three other mesocosms within the same metacommunity. For the low dispersal treatment, 1 L of water was poured reciprocally into each of the other mesocosms in the metacommunity and the remaining 27 L were poured back into the source mesocosm to ensure that the total disturbance from the dispersal treatment was constant across dispersal levels. In the no dispersal treatment, all water was poured back into the mesocosm from which it was collected. All equipment used in the dispersal treatments was rinsed with tap water in between each dispersal event to minimize unintentional dispersal. Dispersal treatments were repeated once per week.

## Statistical Analysis
All analyses were conducted in R version `r R.Version()$major`.`r R.Version()$minor` [@rcoreteam2022]. We assessed whether the experimental treatments resulted in significant differences in the composition of the zooplankton communities using PERMANOVA based on Bray-Curtis dissimilarities using the vegan package [@oksanen2011]. To assess whether the influence of the treatments varied through time, we also included an interactive effect of sampling week in the model. To account for the non-independence of the mesocosms across time and within metacommunities, we included mesocosm and metacommunity identities as nested strata. Then, to determine post-hoc how the treatment effects differed by week, we ran PERMANOVA separately on data from each sampling week, with metacommunity as a strata. If temperature is the primary determinant of composition (H1), we expected to see significant temperature effects on composition across all sampling weeks. If dispersal either facilitates species sorting or homogenizes the metacommunity (H2), we expected to see a significant interaction between temperature and dispersal on community composition. If biotic interactions impede species sorting or result in successional dynamics (H3), we expect to see the influence of temperature on composition depend on time. We visualized compositional differences across treatments and time using non-metric multidimensional scaling (NMDS) based on Bray-Curtis dissimilarity and relative abundance data.

We then used the Hierarchical Modelling of Species Communities (HMSC) package [@tikhonov2021] to estimate the taxon specific responses to our experimental treatments that resulted in the compositional differences that we identified using PERMANOVA. HMSC is a Bayesian generalized mixed effects model that allowed us to estimate the responses of the individual species, based on changes in abundance, to our fixed effects---temperature, dispersal, and time (sampling week)---while accounting for mesocosm and metacommunity as random effects [@ovaskainen2020]. We modelled the zooplankton community as count data using a Poisson distribution and a log-link function.

We estimated the temperature--abundance relationships of the species as quadratic functions, as we assume that each species has a response to temperature such that their abundances peak at intermediate temperatures. For this, the abundance of each species was linked to the water temperature of the mesocosm from which it was sampled, using the average temperature from the week preceding the day of sampling. Thus, the temperature associated with each mesocosm differed across the three sampling dates, depending on both the ambient temperature and the wattage of the heater present (the experimental heat treatment). If temperature is the primary determinant of community composition (H1), we expect to see significant temperature responses that vary between species that lead to predictable turnover in community composition across the spatial temperature gradient between mesocosms as well as through time within the individual mesocosms.

@@

Because our experimental design allowed us to control for environmental differences other than those due to temperature, we assume that additional variation in species' abundances that is not associated with our fixed effects for temperature and dispersal is likely due to biotic interactions (but see discussion for caveats). We quantified this additional variation in three ways. First, we included time (sampling week) as a linear fixed effect in our model to account for temporal changes in species' abundances that were not associated with temperature of dispersal. We elected not to include time as a random or quadratic effect because the temperature was the highest in the middle of the three sampling weeks and so temperature and time would have been confounded in our model. Thus, we assumed that species' abundances that peaked or dipped in the middle sampling week compared to the first and last sample weeks were associated with temperature. Second, mesocosm identity was included as a random effect, allowing us to estimate whether species showed non-random patterns of co-occurrence within individual mesocosms that were not associated with temperature or dispersal. Finally, metacommunity identity was also included as a random effect, allowing us to estimate whether species showed non-random spatial patterns of co-occurrence at the metacommunity scale that were not associated with temperature or dispersal. Thus, if biotic interactions impede species sorting or result in successional dynamics (H3) then we expect to see significant time-abundance relationships and non-random species co-occurrence patterns at either the mesocosm or metacommunity scale.

@@

We assessed the influence of our fixed and random effects using the variance partitioning method within the HMSC package [@tikhonov2021]. Because this method does not account for covariance between explanatory variables [@ovaskainen2020] it cannot properly distinguish the variance explained by temperature and dispersal from that which is explained by their interaction. Thus, we grouped temperature, dispersal, and their interaction together and estimated how much variance they explained in combination. We assessed the influence of our fixed effects on individual species based on whether they showed a positive or negative response. Significance was assessed based on whether or not the 95% credible interval overlapped zero. We also used the fitted model to estimate how species abundances should vary across a temperature range observed in the sampling weeks (19--29 \degree C), and how these response curves vary depending on the three levels of the dispersal treatment. We assessed whether species showed positive or negative co-occurrence patterns, after accounting for the fixed effects. Again we assessed significance based on whether or not the 95% credible interval overlapped zero.

@@

The temperature treatment resulted in compositional differences, but counter to H1 and consistent with H3, these differences were not constant across time (temperature x time; $F_{1,143}$ = `r round(perman$F[5],2)`, p = `r perman[[5]][5]`). Our post-hoc PERMANOVA indicated that temperature had significant effects on composition in week four (i.e, prior to the heatwave; $F_{1,47}$ = `r round(perman4$F[1],2)`, p = `r perman4[[5]][1]`), but not in week 8 (i.e., during the heatwave; $F_{1,47}$ = `r round(perman8$F[1],2)`, p = `r perman8[[5]][1]`) or 12 (i.e., after the heatwave; $F_{1,47}$ = `r round(perman12$F[1],2)`, p = `r perman12[[5]][1]`). This can be seen in the NMDS plots (Figure \@ref(fig:nmds)a), where the communities were clearly structured across the temperature treatment gradient in week 4. However, the heatwave caused the communities to converge on the same compositional state regardless of temperature by week 8. The communities diverged in composition following the heatwave (i.e., week 12), but these communities were no longer differentiated based on temperature as they were in week 4.

Consistent with H2, the dispersal treatments altered the influence of temperature on composition ($F_{2,47}$ = `r round(perman$F[4],2)`, p = `r perman[[5]][4]`), and the effect of dispersal nor its interaction with temperature did not depend on time (dispersal x week, $F_{2,47}$ = `r round(perman$F[6],2)`, p = `r perman[[5]][6]`; dispersal x temperature x week, $F_{2,47}$ = `r round(perman$F[7],2)`, p = `r perman[[5]][7]`). This interactive effect of dispersal and temperature can be seen in the NMDS plots where the composition of mesocosms with dispersal tended to be more similar to the warmer mesocosms (Figure \@ref(fig:nmds)b).

@@

Consistent with H1, all species showed strong and varied temperature--abundance relationships as is necessary for temperature driven species sorting (Figure \@ref(fig:tempcurves)). The taxa differed in the range of temperatures where they were most abundant (Figure \@ref(fig:tempcurves)) as indicated by the varied varied linear temperature coefficients (Figure \@ref(fig:HMSCfig)b).  Daphnia and Calanoida had the coldest optima at 22.5 and 24.5 \degree C respectively and Chaoborus, Cyclopoida, and Diaphanosoma had the warmest optima, which were all above 30 \degree C. In addition, all taxa were estimated to have unimodal temperature--abundance relationships (Figure \@ref(fig:tempcurves)) as indicated by the negative quadratic temperature coefficients (Figure \@ref(fig:HMSCfig)b). These negative quadratic responses were all significant with the exception of Cyclopoida.

As expected under H2, dispersal did alter the realized effects of temperature on population abundance. The most common effect was that dispersal weakened the temperature--abundance relationships, as indicated by the positive dispersal by temperature$^2$ coefficients (Figure \@ref(fig:HMSCfig)b, Figure \@ref(fig:zoopgrid)). Contrary to H2, where we expected this to occur under high but not low dispersal, this weakening of the temperature--abundance relationships was equally common under both low and high dispersal, with five taxa showing significant effects in each dispersal rate. The less common effect of dispersal, exhibited by only two taxa under each dispersal rate, was a strengthening of the temperature-abundance relationship (i.e, negative dispersal by temperature$^2$ coefficients; Figure \@ref(fig:HMSCfig)b). Dispersal had more variable effects on the linear temperature responses, with both positive and negative dispersal temperature interactions being common at both dispersal rate (Figure \@ref(fig:HMSCfig)b). This indicates that dispersal was altering the temperatures at which abundances were highest, but not in a consistent way. Low dispersal caused Scapholeberis to  decreased and Chironomidae to increase in abundance but otherwise dispersal had limited direct inflence on taxon abundances (Figure \@ref(fig:HMSCfig)b).

>>>>

\newpage
# Abstract
Despite the key role of biotic interactions in structuring ecological communities, their influence is often overlooked in predictions of how communities respond to abiotic environmental change. Here, we present an experiment that tests hypotheses based on metacommunity theory about how abiotic responses, biotic interactions and dispersal jointly determine the response of ecological communities to abiotic perturbations. We established experimental zooplankton metacommunities that included a temperature gradient among local communities in each metacommunity. Metacommunities were exposed to one of three dispersal rate treatments. After four weeks, differences in community composition were explained by temperature and this spatial structure was strongest in metacommunities with lower rates of dispersal. A midsummer natural heatwave strongly shifted all communities toward heat tolerant taxa and when conditions cooled, communities diverged into compositional states that were no longer explained by experimental temperature and dispersal treatments. Although our experimental design did not allow us to directly assess the influence of biotic interactions, these results are consistent with the hypothesis that biotic interactions resulted in post-heatwave successional dynamics that prevented the reestablishment of pre-heatwave signatures of temperature on community composition. These results highlight the need to better account for the role of biotic interactions in the theories that guide our expectations for how ecological communities will respond to future abiotic environmental change.

@@

As the climate changes, extreme weather events, such as heatwaves, droughts, and storms, are increasing in frequency and intensity [@ipcc2021], placing ecological communities under considerable stress [@hesketh2023]. To design strategies aimed at mitigating the impacts of climate change on natural ecosystems, we require a sound understanding of the processes that determine how species and ecological communities respond to abiotic change. According to metacommunity theory, the response of communities to abiotic environmental change should depend on the interplay between three general processes: density-independent population responses to abiotic conditions, density-dependent biotic interactions (i.e., interactions between species), and dispersal between communities [@thompson2020]. There is considerable experimental evidence that each of these processes can play a key role in governing community responses [summarized in @leibold2018]. Yet, we still lack clear empirical examples to help understand how abiotic responses, biotic interactions, and spatial processes interactively determine metacommunity dynamics when conditions change.

@@

Metacommunity theory [@loreau2003; @thompson2020] and evidence from a series of experiments [@thompson2012; @symons2013; @limberger2019] suggest that dispersal can play an important role in facilitating community recovery following periods of extreme abiotic conditions. Dispersal can allow species to colonize and re-establish in habitats in which they have been extirpated by harsh abiotic conditions, acting to maintain biodiversity and ecosystem functioning [@loreau2003; @thompson2020]. Thus, dispersal promotes species sorting---that is, the process by which species distributions and abundances become matched to the local conditions across space and time [@leibold2004]. Dispersal can also introduce species that are more tolerant to extreme conditions and if these species perform similar roles to the species they replace, they can compensate for their loss [@thompson2012; @symons2013; @limberger2019]. Colonizing species, however, can also greatly alter the composition and functioning of ecosystems, often with profoundly negative impacts, as highlighted by numerous examples of invasive species [@ricciardi2013]. In addition to facilitating colonization, high dispersal rates can also homogenize communities across space, even when abiotic conditions differ in ways that would lead to community differences with lower dispersal rates, a process known as mass effects [@mouquet2003; @leibold2004]. If rates are high enough, dispersal can maintain populations in environments in which they would otherwise not persist, or in which they would otherwise persist at lower abundances [@holt1985]. Therefore, abiotic conditions are expected to have less influence on the composition of communities when dispersal rates are high, and some species may persist in habitats with less than suitable environments, due to the homogenizing effects of dispersal [@loreau2003]. The overall expectation from metacommunity theory is that intermediate rates of dispersal should facilitate colonization and potentially community recovery from abiotic change via species sorting, and high rates of dispersal should reduce the impact of abiotic change on ecological communities through mass effects [@loreau2003; @thompson2020].

Biotic interactions are expected to cause community assembly and composition patterns to diverge from expectations based on abiotic conditions. The abundance and distribution of species depends on abiotic conditions as well as on the presence and abundance of other species that affect population sizes through competition, facilitation, or prediation [@hutchinson1957]. Competition and predation, for example, may prevent a population from establishing and growing when abiotic conditions would otherwise be suitable [@brown2014]. Thus, biotic interactions are a primary reason why communities that experience the same abiotic conditions may differ in composition [@davis1998; @ives2004; @thompson2021]. Despite this, most classic metacommunity theory [e.g., @loreau2003; @leibold2004] is expressed in simple terms (competition is the only species interaction) and employs the assumption that the strength of competition is equal amongst all individuals and species. Thus, much of this body of theory assumes that community composition is purely determined by abiotic conditions and dispersal. Metacommunity models that explicitly incorporate differences in strength of competition reveal the importance of biotic interactions in structuring communities, and make the prediction that biotic interactions reduce species sorting and lead to non-random patterns of species co-occurrence across space and time that are unexplained by local abiotic conditions [@thompson2020]. When communities experience periods of extreme abiotic conditions, such as during a heatwave, biotic interactions may prevent the community from returning to its initial compositional state [@davis1998; @thompson2021].

While the influence of dispersal and changing abiotic conditions, including temperature, on community composition in spatially structured landscapes can be easily assessed through experimental manipulation [e.g., @symons2013; @limberger2019], experimentally assessing the role of biotic interactions in such systems is more challenging. In microcosms, the role of biotic interactions in shaping community composition across abiotic gradients has been experimentally tested by comparing the response of species to abiotic change in monoculture vs. mixture [@davis1998; @thompson2021]. Because, in this experimental design, each species requires its own set of replicate experimental units [e.g., 216 microcosms in @thompson2021], scaling this approach up to larger scales such as mesocosms is prohibitive. Without being able to experimentally manipulate species interactions, an alternative is to control abiotic conditions and dispersal and then use statistical models to test whether community composition patterns can be explained by these the experimental treatments; if not, the alternative explanation is biotic interactions. Examples of such results would be positive or negative species co-occurrence patterns that are not explained by abiotic treatments as well as temporal shifts in community composition that are not explained by changes in abiotic conditions [@ovaskainen2020].

Here we present the results of a replicated pond zooplankton metacommunity experiment in which we tested how abiotic responses, dispersal, and biotic interactions shape the response and recovery of the community to a mid-summer heatwave. We test the following three hypotheses: H1 -- Community composition is primarily determined by abiotic conditions (here, temperature). If true, we predict that species abundances would vary strongly across spatial and temporal temperature gradients and community composition would be explained by temperature. H2 -- Dispersal modifies the effects of abiotic conditions on community composition. From this hypothesis, we predict that the abiotic environment explains variation in community composition when dispersal occurs at moderate rates, but not when dispersal rates are high. H3 -- Biotic interactions can lead to composition patterns not explained by species sorting. If true, we predict that community composition will vary among local communities such that there are negative and positive co-occurrence patterns over space and time that cannot be explained by abiotic conditions.

@@

## Experimental design
We tested hypotheses 1--3 using an outdoor factorial mesocosm experiment with experimental treatments at the local and metacommuniy scales (Figure 1b). Each metacommunity consisted of four local communities; local communities differed in temperature (4 levels), and metacommunities differed in dispersal rates (low, medium and high). Dispersal treaments were replicated four times, for a total of 48 local communities organized into 12 metacommunities. This design allowed us to test the effects of an abiotic variable on community composition in the presence of three different dispersal rates, and if neither of these factors is important but significant variation in composition were to emerge, we can infer biotic interactions as a possible explanation.

Local communities were assembled in 1136 L aquatic mesocosms (Rubbermaid Stock Tanks, Rubbermaid USA). Within each metacommunity (a set of four adjacent mesocosms), we established a spatial temperature gradient using aquarium heaters (JÄGER TruTsemp, EHEIM GmbH & Co KG, Germany) of different wattages: 100W, 200W, 300W, in addition to a treatment with no heater (i.e., 0W). Thermostats on the heaters were set to their maximum level, 37 \degree C, to ensure that they remained on continuously over the course of the entire experiment, delivering a consistent amount of heat. Mesocosms receiving heaters were placed on 5.08 cm thick rigid panel Styrofoam insulation to increase heat retention. The heaters brought the mesocosms a relatively consistent amount above ambient temperature, resulting in spatial differences in temperature that were maintained as the ambient temperatures fluctuated diurnally and over the course of the 12-week experiment (Figure \@ref(fig:temperature)a). Details on the establishment of the biological communities and the sampling protocol are provided in the Supplementary Materials. The experiment started on June 14, 2018, after phytoplankton, zooplankton and benthos had been added, when the dispersal treatment was applied for the first time. The experiment lasted 12 weeks, ending on September 6, 2018. Samples for estimating zooplankton abundance and composition were collected on week four, eight, and at the end (week 12) of the experiment.

Dispersal---none, low, or high---was applied at the metacommunity level through reciprocal exchange of water and biota. For all dispersal rates, 3 x 10 L of water and biota were collected from each mesocosm using a 3L Van Dorn integrated sampler and placed into 3 - 19 L buckets. For the high dispersal treatment, one of these buckets was poured reciprocally into each of the three other mesocosms within the same metacommunity. For the low dispersal treatment, 1 L of water was poured reciprocally into each of the other mesocosms in the metacommunity and the remaining 27 L were poured back into the source mesocosm to ensure that the total disturbance from the dispersal treatment was constant across dispersal rates. In the no dispersal treatment, all water was poured back into the mesocosm from which it was collected. All equipment used in the dispersal treatments was rinsed with tap water in between each dispersal event to minimize unintentional dispersal. Dispersal treatments were repeated once per week.

## Statistical Analysis
All analyses were conducted in R version `r R.Version()$major`.`r R.Version()$minor` [@rcoreteam2022]. We tested whether the experimental treatments caused differences in the composition of the zooplankton communities using PERMANOVA based on Bray-Curtis dissimilarities using the vegan package [@oksanen2011]. To test whether the influence of the treatments varied through time, we included an interactive effect of sampling week in the model. To account for the non-independence of repeated sampling from the same mesocosms over time and also their spatial proximity within metacommunities, we included mesocosm and metacommunity identities as nested strata (i.e., grouping factors). Then, to describe post-hoc how treatment effects differed by week, we ran PERMANOVA separately on data from each sampling week, with metacommunity as a stratum. If temperature were the primary determinant of composition (H1) and biotic interactions were not a strong influence, we would expect to see significant temperature effects on composition across all sampling weeks. If dispersal either facilitates species sorting or homogenizes composition across communities within a metacommunity (H2), we expect to see a significant interaction between temperature and dispersal on community composition. If biotic interactions impede species sorting (H3), we expect to see the influence of temperature on composition depend on time. We visualized compositional differences across treatments and time using non-metric multidimensional scaling (NMDS) based on Bray-Curtis dissimilarity and relative abundance data.

We then used the Hierarchical Modelling of Species Communities (HMSC) package [@tikhonov2021a] to estimate the taxon specific responses to our experimental treatments. This analysis allowed us to assess which species were responding to treatments in ways that resulted in the compositional differences that we identified using PERMANOVA. HMSC is a Bayesian generalized mixed effects model that allowed us to estimate the responses of the individual species, based on changes in abundance, to our fixed effects---temperature, dispersal, and time (sampling week)---while accounting for mesocosm and metacommunity as random effects [@ovaskainen2020]. We modelled the zooplankton community as count data using a Poisson distribution and a log-link function.

We estimated the temperature--abundance relationships for the zooplankton species as quadratic functions because we assume that each species has a unimodal response to temperature. For this, the abundance of each species was modelled as a function of the water temperature of the mesocosm from which it was sampled, using the average temperature from the week preceding the day of sampling. If temperature is the primary determinant of community composition (H1), we expected to see temperature-abundance coefficients that vary between species and that explain turnover in community composition across the spatial temperature gradient between mesocosms as well as through time within the individual mesocosms.

@@

Because our experimental design allowed us to control for abiotic differences other than those due to temperature, we infer that additional variation in species' abundances that is not associated with our fixed effects for temperature and dispersal is likely due to biotic interactions (but see discussion for caveats). We modelled variation that was unexplained by our treatments in three ways. First, we included time (sampling week) as a linear fixed effect in our model to account for temporal changes in species' abundances that were not associated with temperature of dispersal. We elected not to include time as a random or quadratic effect because the temperature was the highest in the middle of the three sampling weeks and so temperature and time would have been confounded in our model. Thus, we assumed that species' abundances that peaked or dipped in the middle sampling week compared to the first and last sample weeks were associated with temperature. Second, mesocosm identity was included as a random effect, allowing us to estimate whether species showed non-random patterns of co-occurrence within individual mesocosms that were not associated with temperature or dispersal. Finally, metacommunity identity was also included as a random effect, allowing us to estimate whether species showed non-random spatial patterns of co-occurrence at the metacommunity scale that were not associated with temperature or dispersal. Thus, if biotic interactions impede species sorting or result in successional dynamics (H3) then we expect to see strong time-abundance relationships and non-random species co-occurrence patterns at either the mesocosm or metacommunity scale.

@@

We assessed the influence of our fixed and random effects using the variance partitioning method within the HMSC package [@tikhonov2021a]. Because this method does not account for covariance between explanatory variables [@ovaskainen2020] it cannot properly distinguish the variance explained by temperature and dispersal from that which is explained by their interaction. Thus, we grouped temperature, dispersal, and their interaction together and estimated how much variance they explained in combination. We assessed the influence of our fixed effects on individual species based on whether they showed a positive or negative response and whether or not the 95% credible interval overlapped zero. We also used the fitted model to estimate how species abundances should vary across a temperature range observed in the sampling weeks (19--29 \degree C), and how these response curves vary depending on the three rates of the dispersal treatment. We assessed whether species showed positive or negative co-occurrence patterns, after accounting for the fixed effects. Again we assessed whether or not the 95% credible interval overlapped zero.

@@

Our experimental treatments resulted in a mean temperature difference of 2.95 \degree C ( $\pm 0.41$ \degree C SD) between the mesocosms without heaters and the mesocosms that received 300 W heaters (Figure \@ref(fig:temperature)). Ambient temperatures were moderate over the first five weeks of the experiment (until mid-July), with an average mean (max) daily temperature in the unheated mesocosms of `r all_temperature %>% filter(time < 6, wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time < 6, wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C; Figure \@ref(fig:temperature)). The warmest period of the experiment occurred from weeks six to eight (late July and early August), which resulted in an averaged mean (max) daily ambient temperatures of `r all_temperature %>% filter(time %in% c(6:8), wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% c(6:8), wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) and `r all_temperature %>% filter(time %in% c(6:8), wattage == 300) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% c(6:8), wattage == 300) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) in the 300W treatments. Ambient temperatures dropped following this warm period, continuing to decline over the final four weeks of the experiment, so that mean (max) daily temperatures in the unheated mesocosms averaged `r all_temperature %>% filter(time %in% max(time), wattage == 0) %>% pull(temp_mean) %>% mean() %>% round(1)` \degree C (`r all_temperature %>% filter(time %in% max(time), wattage == 0) %>% pull(temp_max) %>% mean() %>% round(1)` \degree C) in the final week.

The temperature treatment resulted in compositional differences, but counter to H1 and consistent with H3, these differences were not constant across time (temperature x time; $F_{1,143}$ = `r round(perman$F[5],2)`, p = `r perman[[5]][5]`). Our post-hoc PERMANOVA indicated that temperature had significant effects on composition in week four (i.e, prior to the heatwave; $F_{1,47}$ = `r round(perman4$F[1],2)`, p = `r perman4[[5]][1]`), but not in week 8 (i.e., during the heatwave; $F_{1,47}$ = `r round(perman8$F[1],2)`, p = `r perman8[[5]][1]`) or 12 (i.e., after the heatwave; $F_{1,47}$ = `r round(perman12$F[1],2)`, p = `r perman12[[5]][1]`). This can be seen in the NMDS plots (Figure \@ref(fig:nmds)a), where the communities were clearly structured across the temperature treatment gradient in week 4. However, the heatwave caused the communities to converge on the same compositional state, regardless of temperature, by week 8. The communities diverged in composition following the heatwave (i.e., week 12), but these communities were no longer differentiated based on temperature as they were in week 4.

Consistent with H2, the dispersal treatments altered the influence of temperature on composition ($F_{2,47}$ = `r round(perman$F[4],2)`, p = `r perman[[5]][4]`), and the effect of dispersal nor its interaction with temperature did not depend on time (dispersal x week, $F_{2,47}$ = `r round(perman$F[6],2)`, p = `r perman[[5]][6]`; dispersal x temperature x week, $F_{2,47}$ = `r round(perman$F[7],2)`, p = `r perman[[5]][7]`). This interactive effect of dispersal and temperature is clear in the comparison of composition among mesocosms in metacommunities in which communities were connected by dispersal; the composition of the individual communities in these these metacommunities tended to be similar to the composition of the warmest mesocosms in the absence of dispersal (Figure \@ref(fig:nmds)b).

@@

Consistent with H1, all species showed strong and varied temperature--abundance relationships as is necessary for temperature driven species sorting (Figure \@ref(fig:tempcurves)). Taxa differed in the range of temperatures at which they were most abundant (Figure \@ref(fig:tempcurves)) as indicated by the varied varied linear temperature coefficients (Figure \@ref(fig:HMSCfig)b).  Daphnia and Calanoida had the coldest optima at 22.5 and 24.5 \degree C respectively and Chaoborus, Cyclopoida, and Diaphanosoma had the warmest optima, which were all above 30 \degree C. In addition, all taxa were estimated to have unimodal temperature--abundance relationships (Figure \@ref(fig:tempcurves)) as indicated by the negative quadratic temperature coefficients (Figure \@ref(fig:HMSCfig)b). The 95% credible intervals of these negative quadratic responses did not overlap zero, with the exception of Cyclopoida.
